Design and Implementation of Probabilistic Programming Language Anglican

### 2016-08-18 paper/paper.tex
preparing for the next stage

\newcommand*{\dt}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf DT:} 
        {\rm #1}\end{minipage}}}
\newcommand*{\fw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf FW:} 
        {\rm #1}\end{minipage}}}
\newcommand*{\hy}[1]{}
% {\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf HY:} 
%        {\rm #1}\end{minipage}}}
\newcommand*{\jw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf JW:} 
        {\rm #1}\end{minipage}}}

@@

Probabilistic programming systems \cite{GMR+08,MSP14,WVM14,GS15}
represent generative models as programs written in a specialized
language that provides syntax for the definition and conditioning of
random variables. The code for such models is generally concise,
modular, and easy to modify or extend. Typically inference can be
performed for any probabilistic program using one or more generic
inference techniques provided by the system backend, such as
Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
\cite{SDT04}, expectation propagation \cite{MWG+10}, and extensions
of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
% Although these generic techniques are not always as statistically
% efficient as techniques that take advantage of model-specific
% optimizations, probabilistic programming makes it easier to optimize
% models for a specific application in a manner that is efficient in
% terms of the dimensionality of its latent variables.
\cbstart
These generic techniques may be less statistically efficient
than techniques tailored to a specific model, however
probabilistic programming facilitates simpler implementations of
models making the inference inherently faster.
\cbend

%\hy{Minor point. I got slightly confused by the last sentence. What does
%``efficient in terms of the dimensionality ..'' mean?}

@@

An implementation of Anglican must therefore address three issues:
\begin{itemize}
    \item the Clojure syntax to introduce probabilistic Anglican
        code inside Clojure modules;
    \item source-to-source transformation of Anglican programs
        into Clojure, so that probabilistic execution becomes
        possible;
    \item algorithms which run Clojure code, obtained by
        transforming Anglican programs, according to the
        probabilistic operational semantics.
\end{itemize}
\cbstart
Execution of probabilistic programs by inference algorithms is
different from execution of deterministic programs.  A
probabilistic program is executed multiple times, often hundreds
of thousands or even millions of times for a single inference
task. Random choices may affect which parts of the program code
are executed and how often. Many inference algorithms require
re-running the program multiple times partially, from a certain
point on.  Different executions may employ different random
choices, however for efficient inference a correspondence
between random choices in different executions should be
maintained. These are just some of the challenges which were
faced and solved during development of Anglican.
\cbend

@@

\hy{Great introduction. It very nicely explains why Anglican
is born and how it is implemented. If I act as a really picky
reviewer or a reader, I would also expect the introduction
to help me grasp in a high level what new insights or scientific problems 
this development of Anglican have resulted in or solved so far.}

@@

Anglican can be mostly implemented as a regular programming
language, except for the handling of these checkpoints.
Depending on the \textit{inference algorithm}, \texttt{sample}
and \texttt{observe} may result in implicit input/output
operations and control changes. For example, \texttt{observe} in
particle filtering inference algorithms \cite{WVM14} is a non-deterministic
control statement at which a particle can be either replicated
or terminated. Similarly, in Metropolis-Hastings \cite{WSG11},
\texttt{sample} is both an input
\cbstart
statement which `reads' values from a random source,
\cbend
and a non-deterministic control statement
(with delayed effect), eventually affecting acceptance or
rejection of a sample.

%\hy{What do you mean by input in ``both an input..''?}

@@

\hy{I think that the concepts like multimethod and protocol
are specific to Clojure, and they are not well-known among PL
people. I tried to rephase the text so that people can follow
it even when they don't know these concepts. I didn't know
``form'' before, but using it is perhaps ok because it is a
standard terminology for Lisp programmers.}

@@

\cbstart
Compilation of Anglican into Clojure is built around a variant of CPS
transformation. In a basic CPS-transformed program, a
continuation receives a single argument --- the computed value.
In Anglican, there are two flows of computation going in
parallel: values are computed by functional code, and,
at the same time, the state of the probabilistic program,
used by inference algorithms, is updated by 
probabilistic forms. Because of that, in Anglican a continuation
accepts  \textit{two} arguments:
\begin{itemize}
    \item the computed value;
    \item the internal state, bound to the local variable
        \texttt{\$state} in every lexical scope.
\end{itemize}

@@

\subsection {The State}
\cbend

@@

\cbstart
So far we described the basics of Anglican compilation to
Clojure. The described approaches and techniques are important
for grasping the language implementation but relatively well-known.
In the rest of the section we focus on challenges we met and
resolved while implementing Anglican, as well as on
implementation of unique features of Anglican as a probabilistic
programming language.
\cbend

@@

There are two proper probabilistic forms turning Anglican into a
probabilistic programming language --- \texttt{sample} and
\texttt{observe}. Their purpose is to interrupt deterministic
computation and transfer control to the inference algorithm.
Practically, this is achieved through returning
\cbstart
\textbf{checkpoints} --- Clojure values of the corresponding
types (\texttt{anglican.trap.sample} or
\texttt{anglican.trap.observe}). The types contain fields
\cbend
specific to each form, as well as the continuation; calling the
continuation resumes the computation. Checkpoints expose the
program state to the inference algorithm, and the updated state
is re-injected into the computation when the continuation is
called: 
\begin{lstlisting}[style=default]
=> (cps-of-expression '(sample dist) 'cont)
(->sample dist cont $state)
=> (cps-of-expression '(observe dist val) 'cont)
(->observe dist val cont $state)
\end{lstlisting}

@@

\cbstart
\paragraph{Memoization}
The author of a probabilistic model might want to draw a
feature of an entity randomly, but to retain the same drawn
value for the same entity during a single run of the probabilistic
program. For example, a person may have brown or green eyes
with some probability, but the \textit{same} person will
always have the same eye color. This can be achieved 
memoized functions. In Anglican, one might write:
\begin{lstlisting}[style=default]
(let [eye-color (mem (fn [person] 
                       (sample 
                         (categorical
                           ['brown 0.5]
                           ['green 0.5]))))]
  (if (not= (eye-color 'bill) (eye-color 'john))
    (predict (eye-color 'bill))
    (predict (eye-color 'john))))
\end{lstlisting}
\cbend
Memoization is often implemented on top of a mutable
dictionary, where the key is the argument list and the value is
the returned value. However, there are no mutable data
structures in a probabilistic program, hence \texttt{mem}'s
memory is stored as a nested dictionary in the program state
(function \texttt{mem-cps}).  Every memoized function gets a
unique automatically generated identifier. Each time a memoized
function is called, one of two continuations is chosen,
depending on whether the same function (a function with the same
identifier) was previously called in the same run of the
probabilistic program with the same arguments. If the memoized
result is available, the continuation of the memoized function
call is immediately called with the stored result. Otherwise,
the argument of \texttt{mem} is called with a continuation which
first creates an updated state with the memoized result, and
then calls the `outer' continuation with the result and the
updated state:
\begin{lstlisting}[style=default]
=> (mem-cps '(foo))
(let [M (gensym "M")]
  (fn mem23623 [C $state & P]
    (if (in-mem? $state M P)
      ;; previously memoized result
      (fn []
        (C (get-mem $state M P) $state))
      ;; new computation
      (clojure.core/apply
        foo
        ;; memoize result in continuation
        (fn [V $state]
          (fn [] (C V (set-mem $state M P V))))
        $state
        P)))))
\end{lstlisting}

@@

Macro \texttt{defdist} should be used to define distributions.
\cbstart
\texttt{defdist} takes care of defining a separate type
\cbend
for every distribution so that multimethods can be dispatched on
distribution types when needed, e.g. for custom proposals. The
Bernoulli distribution could be defined as follows:
\begin{lstlisting}[style=default]
(defdist bernoulli
  "Bernoulli distribution"
  [p]
  (sample* [this] (if (< (rand) p) 1 0))
  (observe* [this value]
    (Math/log (case value
                1 p
                0 (- 1. p)
                0.))))
\end{lstlisting}

>>>>

\newcommand*{\dt}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf DT:} 
        {\rm #1}\end{minipage}}}
\newcommand*{\fw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf FW:} 
        {\rm #1}\end{minipage}}}
\newcommand*{\hy}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf HY:} 
        {\rm #1}\end{minipage}}}
\newcommand*{\jw}[1]{\smallskip\noindent\fbox{\begin{minipage}{\linewidth}{\bf JW:} 
        {\rm #1}\end{minipage}}}

@@

Probabilistic programming systems \cite{GMR+08,MSP14,WVM14,GS15}
represent generative models as programs written in a specialized
language that provides syntax for the definition and conditioning of
random variables. The code for such models is generally concise,
modular, and easy to modify or extend. Typically inference can be
performed for any probabilistic program using one or more generic
inference techniques provided by the system backend, such as
Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
\cite{SDT04}, expectation propagation \cite{MWG+10}, and extensions
of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
% Although these generic techniques are not always as statistically
% efficient as techniques that take advantage of model-specific
% optimizations, probabilistic programming makes it easier to optimize
% models for a specific application in a manner that is efficient in
% terms of the dimensionality of its latent variables.
These generic techniques may be less statistically efficient
than techniques tailored to a specific model, however
probabilistic programming facilitates simpler implementations of
models making the inference inherently faster.

@@

An implementation of Anglican must therefore address three issues:
\begin{itemize}
    \item the Clojure syntax to introduce probabilistic Anglican
        code inside Clojure modules;
    \item source-to-source transformation of Anglican programs
        into Clojure, so that probabilistic execution becomes
        possible;
    \item algorithms which run Clojure code, obtained by
        transforming Anglican programs, according to the
        probabilistic operational semantics.
\end{itemize}
Execution of probabilistic programs by inference algorithms is
different from execution of deterministic programs.  A
probabilistic program is executed multiple times, often hundreds
of thousands or even millions of times for a single inference
task. Random choices may affect which parts of the program code
are executed and how often. Many inference algorithms require
re-running the program multiple times partially, from a certain
point on.  Different executions may employ different random
choices, however for efficient inference a correspondence
between random choices in different executions should be
maintained. These are just some of the challenges which were
faced and solved during development of Anglican.

@@

@@

Anglican can be mostly implemented as a regular programming
language, except for the handling of these checkpoints.
Depending on the \textit{inference algorithm}, \texttt{sample}
and \texttt{observe} may result in implicit input/output
operations and control changes. For example, \texttt{observe} in
particle filtering inference algorithms \cite{WVM14} is a non-deterministic
control statement at which a particle can be either replicated
or terminated. Similarly, in Metropolis-Hastings \cite{WSG11},
\texttt{sample} is both an input
statement which `reads' values from a random source,
and a non-deterministic control statement
(with delayed effect), eventually affecting acceptance or
rejection of a sample.

@@

@@

\dt{I believe I partially addressed the above comment, but need to
think more about it.}

Compilation of Anglican into Clojure is built around a variant of CPS
transformation. In a basic CPS-transformed program, a
continuation receives a single argument --- the computed value.
In Anglican, there are two flows of computation going in
parallel: values are computed by functional code, and,
at the same time, the state of the probabilistic program,
used by inference algorithms, is updated by 
probabilistic forms. Because of that, in Anglican a continuation
accepts  \textit{two} arguments:
\begin{itemize}
    \item the computed value;
    \item the internal state, bound to the local variable
        \texttt{\$state} in every lexical scope.
\end{itemize}

@@

\subsection {The State}

@@

So far we introduced the basics of Anglican compilation to
Clojure. The described approaches and techniques are important
for grasping the language implementation but relatively well-known.
In the rest of the section we focus on challenges we met and
resolved while implementing Anglican, as well as on
implementation of unique features of Anglican as a probabilistic
programming language.

@@

There are two proper probabilistic forms turning Anglican into a
probabilistic programming language --- \texttt{sample} and
\texttt{observe}. Their purpose is to interrupt deterministic
computation and transfer control to the inference algorithm.
Practically, this is achieved through returning
\textbf{checkpoints} --- Clojure values of the corresponding
types (\texttt{anglican.trap.sample} or
\texttt{anglican.trap.observe}). The types contain fields
specific to each form, as well as the continuation; calling the
continuation resumes the computation. Checkpoints expose the
program state to the inference algorithm, and the updated state
is re-injected into the computation when the continuation is
called: 
\begin{lstlisting}[style=default]
=> (cps-of-expression '(sample dist) 'cont)
(->sample dist cont $state)
=> (cps-of-expression '(observe dist val) 'cont)
(->observe dist val cont $state)
\end{lstlisting}

@@

\paragraph{Memoization}
The author of a probabilistic model might want to draw a
feature of an entity randomly, but to retain the same drawn
value for the same entity during a single run of the probabilistic
program. For example, a person may have brown or green eyes
with some probability, but the \textit{same} person will
always have the same eye color. This can be achieved 
memoized functions. In Anglican, one might write:
\begin{lstlisting}[style=default]
(let [eye-color (mem (fn [person] 
                       (sample 
                         (categorical
                           ['brown 0.5]
                           ['green 0.5]))))]
  (if (not= (eye-color 'bill) (eye-color 'john))
    (predict (eye-color 'bill))
    (predict (eye-color 'john))))
\end{lstlisting}

Memoization is often implemented on top of a mutable
dictionary, where the key is the argument list and the value is
the returned value. However, there are no mutable data
structures in a probabilistic program, hence \texttt{mem}'s
memory is stored as a nested dictionary in the program state
(function \texttt{mem-cps}).  Every memoized function gets a
unique automatically generated identifier. Each time a memoized
function is called, one of two continuations is chosen,
depending on whether the same function (a function with the same
identifier) was previously called in the same run of the
probabilistic program with the same arguments. If the memoized
result is available, the continuation of the memoized function
call is immediately called with the stored result. Otherwise,
the argument of \texttt{mem} is called with a continuation which
first creates an updated state with the memoized result, and
then calls the `outer' continuation with the result and the
updated state:
\begin{lstlisting}[style=default]
=> (mem-cps '(foo))
(let [M (gensym "M")]
  (fn mem23623 [C $state & P]
    (if (in-mem? $state M P)
      ;; previously memoized result
      (fn []
        (C (get-mem $state M P) $state))
      ;; new computation
      (clojure.core/apply
        foo
        ;; memoize result in continuation
        (fn [V $state]
          (fn [] (C V (set-mem $state M P V))))
        $state
        P)))))
\end{lstlisting}

@@

Macro \texttt{defdist} should be used to define distributions.
\texttt{defdist} takes care of defining a separate type
for every distribution so that multimethods can be dispatched on
distribution types when needed, e.g. for custom proposals. The
Bernoulli distribution could be defined as follows:
\begin{lstlisting}[style=default]
(defdist bernoulli
  "Bernoulli distribution"
  [p]
  (sample* [this] (if (< (rand) p) 1 0))
  (observe* [this value]
    (Math/log (case value
                1 p
                0 (- 1. p)
                0.))))
\end{lstlisting}
